The tree line ecotone punctuates the elevational turnover of plants, pollinators, and their interactions 

### 2021-03-17 ms_1/ms_1.Rmd
update

Climate-induced range shifts involve more than merely tracking suitable temperature envelopes; when species move, they do so in the context of biological communities and under the influence of more than just temperature, and the outcomes of climate-induced range shifts are thus contingent on processes that cannot be reduced to linear elevational or latitudinal temperature gradients. Bumble bees (Hymentopera: *Bombus* spp.) are cold-adapted species associated with mountain habitats, and they have already exhibited measurable upslope range shifts that are expected to result in habitat loss, novel competitive interactions, and disrupted pollination networks. Interpreting these shifts, however, requires an understanding of the current elevational patterns of bumble bees and the floral mutualists upon which climate change will act. In this study, we analyze bumble bee, flowering plant, and interaction records sampled during 904 standardized transect walks conducted over three consecutive years along an elevational gradient in the German Alps ranging from 640 to 2030 meters above sea level.

We demonstrate that mountain bumble bees and flora exhibit distinct peaks of abundance, are distributed in strongly nonlinear patterns in space and time, and that these patterns propagate from the level of species abundances to that of β-diversity, and from the level of community composition to that of biotic interactions. In particular, we show that the tree line ecotone is a critical threshold of floral resource availability, a distributional interface between high- and low-elevation bumble bees species, and a zone of rapid acceleration of species and interaction β-diversity. The implications of these findings extend beyond the particular case of bumble bees to demonstrate that linear elevational temperature gradients are ecologically punctuated, and the outcomes of climate-induced range shifts with depend on how species and species-interactions change as they cross --- or fail to cross --- thresholds like the tree line ecotone.

*Keywords*: climate, elevation gradient, pollinator, *Bombus*, alpine

@@

The rate of recent climate change is forcing species to adapt ecologically within the constraints of gene pools shaped by cooler climates [@Visser2008-ct]. As temperatures rise, species are expected to track suitable conditions to higher latitudes and higher elevations, and such effects are already measurable across a broad range of taxa [@Lenoir2015-tj]. Species range shifts, however, are not merely abstract translations along axes of latitude and elevation; they are movements *in situ*, involving the advent, extinction, and reorganization of biological interactions [@Tylianakis2008-fi; @Blois2013-dz] as well as the concomitant variation of abiotic conditions other than temperature [@Korner1995-mh; @Hodkinson2005-af]. Thus, when species move along continuous, linear climate gradients, they can experience punctuated and nonlinear ecological transitions. A salient example is the tree line ecotone that marks the elevational transition between forest and grassland. At this relatively discrete band within a continuous elevational temperature gradient, plant communities exhibit rapid species turnover [@Descombes2017-id] and declining diversity [@Becker2007-cg] as abiotic conditions become more extreme away from the shelter of tree canopy [@Slatyer1992-ir]. Thus, species range shifts and the consequences thereof are contingent upon existing patterns of species distributions and biotic interactions.

Bumble bees (Hymenoptera: *Bombus* spp.) are well-studied model organisms in the field of community ecology, but they have more recently become conservation priorities due to well-documented population declines and particular vulnerability to climate change [@Goulson2008-oe; @Soroye2020-zb]. The genus *Bombus* is believed to have arisen in the mountains of Asia during the cooling climate of the Eocene-Oligocene boundary (~34 mya) [@Hines2008-er], and bumble bees remain characteristically cold-adapted species today [@Heinrich1994-cv], with peak abundance and diversity in mountain ranges and northern latitudes where they are often the dominant guild of flower-visiting insects and the principal pollinators of entomophilous flora [@Williams1998-xi; @Goulson2010-lo]. Climate-induced range shifts are expected to be especially pronounced in the mountain habitats that host the core of bumble bee biodiversity, together with their floral mutualists, and upward movement of bumble bee species has already been documented [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc]. Populations of bumble bee species already restricted to the highest elevation zones can be expected to suffer geometric habitat shrinkage and increasing genetic isolation as they move toward mountain peaks; the upward advance of lowland species may introduce novel competitive pressures and disrupt the pollination networks in which bumble bees participate [@Brosi2013-mu; @Ishii2013-sm]; and concurrent elevational [@Lenoir2008-yn] and/or phenological [@Pyke2016-ll] shifts in the floral community could generate mismatches between bumble bee species and their historic floral hosts.

<!-- I either need to beef up the paragraph below or fold it into another one. But I need to get to the point of our study. Which means I need to know what the point is. -->

The outcomes of range shifts in mountain bumble bee and floral communities have obvious significance from a conservation perspective [@Soroye2020-zb], but they also provide a model system in which to study the basic phenomenon of community assembly under climate change, a phenomenon of both historical interest and prospective urgency. As bumble bee species move up mountain slopes, they move in relation to plant species and other bumble bee species

Anticipating, interpreting, and potentially mitigating the effects of ranges shifts, however, hinge upon an understanding the current elevational and temporal patterns of the bumble bee and floral communities that will be acted upon by climate change. In particular, it is important to detect nonlinearities in species-, community-, and interaction-level responses to elevational climate gradients, such as the floristic threshold effect that has been documented at the tree line ecotone [@Descombes2017-id].

⎄
In the present study, we approach this topic at three levels of biological organization. First, we ask how mountain bumble bee species are distributed in elevation and intra-annual time with respect to one another and with respect to their floral hosts. Then, we move from the level of individual species to that of community composition and explore the structure of species turnover (i.e. β-diversity) of bumble bees and flora in response to elevation. Finally, we proceed to the level of species interactions, investigating the turnover of bee-flower interaction partners through elevation. In each of these analyses, we employ flexible modeling techniques capable of capturing nonlinear responses with the goal of identifying where, when, and for which species the impacts of climate-induced range shifts should be expected to be most severe.

@@

We analyzed the abundance of bumble bees and floral resources using separate hierarchical generalized additive models (HGAM) [@Wood2011-pg; @Pedersen2019-kf]. We opted in both models to the collapse temporal variation of abundance within sites and focus on elevational patterns. For bumble bees, we modeled *peak* abundance at each site as a smooth response to elevation, with bumble bee species as a "by" smoothing factor and a random intercept effect. We also included species:year as a random intercept effect to capture annual diffences in species abundance and site as a random intercept effect to avoid overfitting due to pseudoreplication of elevation arising from the measurement of multiple bumble bee species at each site. This construction is a modification of the "type I" model form described in @Pedersen2019-kf. For floral resources, we modeled *mean* flower cover, weighted per bumble be species, as a smooth response to elevation, with bumble bee species and year as "factor-smooths" and site as a random intercept effect (again to avoid overfitting due to pseudoreplication). This construction is a modification of the "type S" model form described in @Pedersen2019-kf.

@@

Our model explained 71.5% of total deviance (adjusted R^2^ = 0.511).

*B. mendax*, *B. monticola*, *B. mucidus*, and *B. pyrenaeus* exhibited a positive response to elevation and peaked in abundance above the tree line (~ 1500 m.s.l.) (**Figure 3**), consistent with their known affinity for high elevation habitats [@Rasmont2010-ci]. The abundance of *B. soroensis* also tended to increase with elevation and peaked above the tree line, but it was also abundant at lower elevations. In contrast, the abundance of *B. pascorum* --- the most abundant species in our study system --- declined steeply with elevation, and a similar pattern was seen for *B. psithyrus*.

while *B. gerstaeckeri* was sparsely distributed at mid-elevation sites.

```{r fig3, echo = FALSE,  fig.cap = paste("Peak bumble bee abundance by elevation and year. Curves shows the conditional GAM predictions excluding the site random effect, and points show original data. Dotted vertical lines indicate the approximate location of the tree line at our study sites. Note different scales of y-axes.")}
knitr::include_graphics("../output/gam_max_bb_abund2.png")
```

```{r fig4, echo = FALSE,  fig.cap = paste("Mean weighted floral abundance (area of flower cover) by elevation and year. Lines shows the conditional GAM predictions excluding the random effects of site and species, and points show original data. The right panel displays the full range of the data while the left panel displays a subset of the range, omitting extreme values of flower cover to highlight the prediction curves.")}
knitr::include_graphics("../output/gam_mean_fl_abund_zoom.png")
```

@@

```{r fig5, echo = FALSE,  fig.cap = paste("Species-level (A) and interaction-level (B, C) plotted against elevational difference. Each point represents the β-diversity (or β-diversity partition) between a pair of sites, and lines represent the overall relationship between β-diversity and elevation difference using binomial regression smooths. Standard errors are not plotted because they would be misleading due to the non-independence inherent to distance matrix regression, but all regressions were significant (p < 0.01). Interaction β-diversity (B, C) is partitioned using Poisot et al.’s (2012) notation: WN = unpartitioned β-diversity, ST = β-diversity due to species turnover, OS =  β-diversity due to interaction rewiring, ST.h =  β-diversity due to species turnover in the higher trophic level (bumble bees), ST.l = β-diversity due to species turnover in the lower trophic level (plants), and ST.lh =  β-diversity due to joint species turnover in higher and lower trophic levels.")}
knitr::include_graphics("../output/beta_stack.png")
```

@@

```{r fig6, echo = FALSE,  fig.cap = paste("GDM splines of species-level and interaction-level β-diversity in response to elevational, controlling for geographic proximity and number of sampling dates. The maximum height of each spline represents a given variable’s partial effect on β-diversity (its effect when covariates are held constant), and the shape of each spline represents to the rate of species turnover as it varies along the corresponding gradient, with steeper parts of the curve indicating regions of the gradient over which species/interaction turnover is more rapid. Dotted vertical lines in the elevation panels indicates the location of the tree line. Within-site comparisons were omitted from GDM analysis to avoid pseudoreplication. Shaded bands depict the uncertainty of each partial effect curve (+/- one standard deviation) based on bootstrapping. Note the differences in scale between the y-axes of each variable.")}
knitr::include_graphics("../output/gdm_stack.png")
```

@@

The core finding of our study, evident in each of our analyses, is that a linear elevation gradient can generate nonlinear biological responses in terms of both abundance and β-diversity, and the latter both at the level of community composition and at the level of species interactions.

Arguably the most salient pattern in our data is the unique importance of the elevation band falling at about 1500 m, which corresponds to the lower edge of the tree line ecotone in the Berchtesgadener Alps [@Kostler1970-ya; @Mayer1970-zn] and specifically that of the elevational transects used in our study. The floristic importance of the tree line ecotone has been noted in previous work [@Pellissier2010-kg; @Descombes2017-id]. By studying the floral community through the lens of foraging bumble bees, however, we provide a functional extension of this floristic pattern, demonstrating that the tree line ecotone comprises (1) a boundary above which floral resource availability sharply declines, (2) a distributional interface between high-elevation bumble bees and their low-elevation or habitat-generalist counterparts, and (3) a zone of rapid acceleration of species and interaction β-diversity. The significance of the tree line ecotone as a threshold of floral resource availability and as a zone of overlapping bumble bee distributions corroborates the recent findings of @Minachilis2020-xd in the Mediterranean system of Mount Olympus, indicating that these patterns are likely to generalize well across mountain ecosystems. Our finding that interaction β-diversity is also punctuated by the tree line ecotone follows naturally from these compositional patterns, and we expect that the β-diversity patterns found in our system will prove similarly generalizable.

The observation that the upslope advancement of alpine tree lines often lags behind climate warming [@Dullinger2004-qu] suggests that high elevation bumble bees could move upslope faster than their floral resources, though the opposite pattern could also occur if the climate-tracking of bumble bees is limited by other factors, such as nest site availability [@Marshall2020-pc]. A somewhat unexpected finding of our study is that high-elevation bumble bee species (*B. mucidus*, *B. monticola*, *B. mendax*, and *B. pyrenaeus*) were associated with floral hosts that peaked in abundance at much lower elevations, largely outside the range of their bumble bee visitors. As the upper boundary of high floral resource abundance, the lower edge of the tree line ecotone likely functions as critical foraging habitat for high-elevation bumble bees. Future warming could force these species farther into resource-poor alpine grasslands while at the same time increasing resource competition at the tree line as dominant low-elevation species, like *B. pascuorum*, extend their ranges upslope.

Our analyses of species and interaction β-diversity reveal a striking contrast between a relative stable bumble bee community and an extremely dynamic floral community. The dominance of total interaction β-diversity by floral species turnover is consistent with the findings of @Simanonok2014-ou in the Rocky Mountains of North America, and it is perhaps a predictable relationship given the exceptional cold-hardiness of bumble bees and the sensitivity of plants to temperature gradients. Nevertheless, it highlights the behavioral challenge that bumble bees have presumably faced for the entirety of their evolutionary history as mountain florivores: how to forage efficiently in a diverse floral community that turns over rapidly in both space and time. Indeed, the remarkable flexibility and intelligence of bumble bee foraging [@Heinrich1979-ep; @Loukola2017-bt], which has been interpreted as a side-effect of their sociality [@Dukas1991-ig], could be interpreted alternatively as an adaptation to precisely this problem of foraging in the context of extreme β-diversity. A fascinating question that could not be answered by our sampling approach is whether mountain bumble bees adaptively forage up- or downslope, as suggested by @Lundberg1980-qk. Bumble bees have large foraging ranges and have been shown both to track resources through space and time [@Devoto2014-ay] and to cross forest matrix to reach patches of foraging habitat [@Mola2020-yc]. Assuming an average slope of 20° and a foraging range of 1 km, a bumble bee could travel up- or downslope by more than 340 m, thus spanning an elevation belt nearly 700 m wide. Such 3-dimensional foraging would enable bumble bees both to exploit the elevational turnover of floral species and to track preferred species through their elevationally staggered phenology, the latter constituting a sort of physiological time travel [@Van_Straalen1983-df].

# CONCLUSIONS

Mountains are unique ecological theaters in which extrinsic temporal climate change interacts with intrinsic elevational climate gradation to alter the spatial and temporal distributions of species and interactions [@Telwala2013-sc; @CaraDonna2014-nq; @Miller-Struttmann2014-wz; @Rafferty2020-aj]. The central finding of our study, that elevation gradients are characterized by nonlinear responses of species, communities, and interaction networks --- punctuated dramatically by the tree line ecotone --- is almost certainly not an idiosyncrasy of bumble bees and their floral hosts but rather a pattern emerging from fundamental physical constraints that should be expected to obtain broadly across the biotic communities of mountains worldwide. Species interactions along tree line ecotones, therefore, deserve special consideration in research and conservation management, since climate-induced range shifts at the tree line ecotone should be expected to have disproportionate ecological consequences relative to range shifts at other elevation bands.

@@

<!-- - What do these findings tell use about climate change and range shifts in mountain ecosystems more generally? -->
<!-- - Do our results suggest any management strategies? -->

<!-- Retreating to higher elevation in response to climatic warming is not a new phenomenon in the evolutionary history of bumble bees, but the rate of modern climate change, together with the interacting effects of human land use, are predicted to [@Soroye2020-zb; @Sirois-Delisle2018-vi]  -->

<!-- though as a matter of applied conservation it is perhaps doubtful that local interventions could forestall for long the consequences of a process occurring at a global scale. -->

>>>>

The structuring of biological communities along mountain slopes is complex, and the elevational range shifts that are now occurring in response to climate change involve more than merely tracking suitable temperature envelopes. When species move, they do so in the context of biological communities, and the outcomes of these movements depend on how and to what extent biotic interactions are reordered. Bumble bees (Hymentopera: *Bombus* spp.) are cold-adapted species associated with mountain habitats, and they have already exhibited measurable upslope range shifts that are expected to result in habitat loss, novel competitive interactions, and the rewiring of pollination networks. Predicting and interpreting these shifts, however, requires an understanding of the current elevational patterns of bumble bees and the floral mutualists that are being acted upon by climate change. In this study, we sampled bumble bees, flowering plants, and their interactions over three consecutive years along an 1400 m elevational gradient in the German Alps. Using nonlinear modeling techniques, we analyze the elevational patterns of this community at three levels of biological organization: species abundance, species β-diversity, and interaction β-diversity. We demonstrate complex, nonlinear responses to elevation at all three levels of organization. In particular, we identify the tree line ecotone as (1) a distributional interface between low/mid- and high-elevation bumble bee species, (2) a threshold above which floral resource availability sharply decreases, and (3) a zone of accelerated turnover of flowering plants and bumble- bee-flower interactions. The implications of these findings extend beyond the particular case of bumble bees to demonstrate that linear elevational temperature gradients are ecologically punctuated, and the outcomes of climate-induced range shifts with depend on how species and species-interactions change as they cross --- or fail to cross --- thresholds like the tree line ecotone.

*Keywords*: climate, elevation gradient, pollinator, *Bombus*, mountain, tree line

@@

<!-- This is a paper about mountains. -->

Mountains are unique ecological theaters in which extrinsic climate dynamics interact with intrinsic climate gradation to shape the distributions of species and biotic interactions [@Telwala2013-sc; @CaraDonna2014-nq; @Miller-Struttmann2014-wz; @Rafferty2020-aj]. The rate of recent climate change is forcing species to adapt ecologically within the constraints of gene pools shaped by cooler climates [@Visser2008-ct]. As temperatures rise, species are expected to track suitable conditions to higher elevations, and such effects are already measurable across a broad range of taxa [@Lenoir2015-tj]. Elevational range shifts, however, are not merely abstract translations along the linear lapse of temperature generated by mountain slopes; they are movements *in situ*, involving the advent, extinction, and reorganization of biological interactions [@Tylianakis2008-fi; @Blois2013-dz] as well as the concomitant variation of abiotic conditions other than temperature [@Korner1995-mh; @Hodkinson2005-af]. Thus, when species move along continuous, linear climate gradients, they can experience punctuated and nonlinear ecological transitions. A salient example is the tree line ecotone that marks the elevational transition between forest and grassland. At this relatively discrete band within a continuous elevational temperature gradient, plant communities exhibit rapid species turnover [@Descombes2017-id] and declining diversity [@Becker2007-cg] as abiotic conditions become more extreme away from the shelter of tree canopy [@Slatyer1992-ir].

Bumble bees (Hymenoptera: *Bombus* spp.) are classic model organisms in the field of community ecology, but they have more recently become conservation priorities due to well-documented population declines and vulnerability to climate change [@Goulson2008-oe; @Soroye2020-zb]. Bumble bees are characteristically cold-adapted species [@Heinrich1994-cv], with peak abundance and diversity in mountain ranges and northern latitudes, where they are often the dominant guild of flower-visiting insects and the principal pollinators of entomophilous flora [@Williams1998-xi; @Goulson2010-lo]. Upslope movement of bumble bee species in response to climate change has already been documented, and this is expected to cause the declines or extinction of some bumble bee species [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc]. From a purely geometric perspective, populations of bumble bee species already restricted to the highest elevation zones can be expected to suffer habitat shrinkage [@Dirnbock2011-jf], possibly accompanied by increasing genetic isolation [though see @Fijen2020-zg]. But the outcomes of elevational range shifts on bumble bees also hinge upon the current elevational distributions of bumble bees and their floral resources that will be acted upon by climate change. The upward advance of lowland bumble bee species, for example, may introduce novel competitive pressures on highland species and rewire the pollination networks in which they participate [@Nagamitsu2010-wi; @Brosi2013-mu; @Ishii2013-sm], and elevational [@Lenoir2008-yn] and/or phenological [@Pyke2016-ll] shifts in the floral community could generate mismatches between bumble bee species and their floral hosts. When range shifts are considered in this complex biotic context, it is especially important to understand nonlinearities or thresholds in species-, community-, and interaction-level responses to elevational climate gradients, such as the floristic threshold effect that has been documented at the tree line ecotone [@Descombes2017-id], since these could produce dramatic ecological consequences that would not be predicted by geometric considerations alone.

The outcomes of range shifts in mountain bumble bee and floral communities have obvious significance from a conservation perspective [@Soroye2020-zb], but they also provide a model system in which to study the basic phenomenon of community assembly under climate change, a phenomenon of both historical interest and prospective urgency. In the present study, set along ~1400 m elevation gradient in the German Alps, we investigate elevational patterns in a community of bumble bees and wildflowers at three levels of biological organization. First, we ask how elevation shapes the abundances of bumble bee species and their floral resources. Then, we move from the level of individual species to that of community composition and explore the elevational structure of species turnover (i.e. β-diversity) in bumble bees and flowering plants. Finally, we advance to the level of species interactions, investigating the turnover of bumble- bee-flower interaction partners through elevation. In each of these analyses, we employ flexible modeling techniques designed to capture nonlinear responses to elevation, with a particular emphasis on the tree line ecotone as a key threshold that structures the biological communities of mountains slopes.

@@

We analyzed the abundance of bumble bees and floral resources using separate hierarchical generalized additive models (HGAMs) fit with the Tweedie distribution family (log link function) [@Wood2011-pg; @Pedersen2019-kf]. We opted in both models to collapse the temporal variation of abundance within sites and focus on elevational patterns. For bumble bees, we modeled *peak* abundance at each site as a smooth response to elevation, with bumble bee species as a smoothing factor and a random intercept term. We also included species:year as a random intercept term because exploratory analysis indicated that species varied differently across years. For floral resources, we modeled *mean* flower cover, weighted per bumble bee species, as a smooth response to elevation, with year and bumble bee species as random intercept terms. Both models are follow the "type I" model form described in @Pedersen2019-kf.

@@

The peak abundance of all bumble bee species responded significantly to elevation (p ≤ 0.04) **(Figures 3, 4)**. *B. mendax*, *B. monticola*, *B. mucidus*, and *B. pyrenaeus* exhibited a positive response to elevation and peaked in abundance above the tree line (~ 1500 m.s.l.), consistent with their known affinity for high elevation habitats [@Rasmont2010-ci]. The abundance of *B. soroeensis* also tended to increase with elevation and peaked above the tree line, but it was also abundant at lower elevations. In contrast, the abundance of *B. hortorum*, *B. pascorum*, and *B. psithyrus* declined with elevation. This effect was especially pronounced in *B. pascuorum*, which was also the most abundant species in our study system overall. *B. gerstaeckeri*, *B. jonellus*, *B. terrestris-lucorum*, and *B. wurflenii* peaked at mid-elevation below the tree line, while *B. pratorum* was unique in exhibiting a sharp peak of abundance directly *at* the tree line.

```{r fig3, echo = FALSE,  fig.cap = paste("Fitted GAM smooths of bumble bee abundance on elevation, faceted by bumble bee species and plotted over original data. Data points are color-coded to indicate whether they represent sites above (red) or below (blue) the tree line. Note the different scales of the y-axes.")}
knitr::include_graphics("../output/gam_bb_abund.png")
```

```{r fig4, echo = FALSE,  fig.cap = paste("Conditional effect of elevation on the peak abundance of each bumble bee species. Effects are plotted on the link scale, and dashed lines depict 95\\% confidence intervals.")}
knitr::include_graphics("../output/gam_bb_abund_grid.png")
```

Mean floral resource abundance likewise responded significantly (p << 0.0001) to elevation for all bumble bee species **(Figures 5, 6)**. *B. mendax*, *B. monticola*, *B. mucidus*, *B. soroeensis*, and *B. wurflenii* exhibited a weakly to strongly bimodal pattern of flora resource abundance, with peaks around 1000 m.s.l. and 1600 m.s.l., the latter peak corresponding closely to the location of the tree line. *B. jonellus* and *B. pyrenaeus* exhibited a unimodal pattern peaking near 1600 m.s.l. The remaning species --- *B. gerstaeckeri*, *B. hortorum*, *B. pascuorum*, *B. pratorum*, *B. psithyrus*, and *B. terrestris-lucorum* experienced linear or quadratic decline in floral resource abundance with increasing elevation. Floral resource abundance was low for all species at elevations above ~1600 m.s.l.

```{r fig5, echo = FALSE,  fig.cap = paste("Fitted GAM smooths of floral abundance on elevation, faceted by bumble bee species and plotted over original data. Data points are color-coded to indicate whether they represent sites above (red) or below (blue) the tree line. Note the different scales of the y-axes.")}
knitr::include_graphics("../output/gam_fl_abund.png")
```

```{r fig6, echo = FALSE,  fig.cap = paste("Conditional effects of elevation on mean flower cover by bumble bee species and year. Effects are plotted on the link scale and dashed lines depict 95\\% confidence intervals.")}
knitr::include_graphics("../output/gam_fl_abund_grid.png")
```

@@

```{r fig7, echo = FALSE,  fig.cap = paste("Species-level (A) and interaction-level (B, C) plotted against elevational difference. Each point represents the β-diversity (or β-diversity partition) between a pair of sites, and lines represent the overall relationship between β-diversity and elevation difference using binomial regression smooths. Standard errors are not plotted because they would be misleading due to the non-independence inherent to distance matrix regression, but all regressions were significant (p < 0.01). Interaction β-diversity (B, C) is partitioned using Poisot et al.’s (2012) notation: WN = unpartitioned β-diversity, ST = β-diversity due to species turnover, OS =  β-diversity due to interaction rewiring, ST.h =  β-diversity due to species turnover in the higher trophic level (bumble bees), ST.l = β-diversity due to species turnover in the lower trophic level (plants), and ST.lh =  β-diversity due to joint species turnover in higher and lower trophic levels.")}
knitr::include_graphics("../output/beta_stack.png")
```

@@

```{r fig8, echo = FALSE,  fig.cap = paste("GDM splines of species-level and interaction-level β-diversity in response to elevational, controlling for geographic proximity and number of sampling dates. The maximum height of each spline represents a given variable’s partial effect on β-diversity (its effect when covariates are held constant), and the shape of each spline represents to the rate of species turnover as it varies along the corresponding gradient, with steeper parts of the curve indicating regions of the gradient over which species/interaction turnover is more rapid. Dotted vertical lines in the elevation panels indicates the location of the tree line. Within-site comparisons were omitted from GDM analysis to avoid pseudoreplication. Shaded bands depict the uncertainty of each partial effect curve (+/- one standard deviation) based on bootstrapping. Note the differences in scale between the y-axes of each variable.")}
knitr::include_graphics("../output/gdm_stack.png")
```

@@

The core finding of our study, evident in each of our analyses, is that a linear elevation gradient can generate complex and nonlinear biological responses in a mountain bumble- bee-wildflower community in terms of both abundance and β-diversity, and the latter both at the level of community composition and at the level of species interactions. As a rule, these responses are punctuated by the tree line ecotone, which constitutes a discrete ecological threshold along the linear elevation gradient of a mountain slope.

We found a clear distinction between bumble bee species with low- or mid-elevation abundance peaks (*B. gerstaeckeri*, *B. hortorum*, *B. jonellus*, *B. pascorum*, *B. pratorum*, *B. psithyrus*, *B. terrestris-lucorum*, *B. wurflenii*) and those with high-elevation abundance peaks (*B. mendax*, *B. monticola*, *B. mucidus*, *B. pyrenaeus*, and *B. soroeensis*). The tree line forms an important zone of overlap at which the declining abundance of low- and mid-elevation species intersects with the rising abundance of high-elevation species. The tree line was also important in the elevational structure of floral resource abundance. For some (mostly high-elevation) bumble bee species, the tree line marked a distinct peak of floral resource availability, often mirrored by a second peak at lower elevation (~1000 m.s.l.). It is notable that, for some of the high-elevation species that exhibited this bimodal pattern of floral resource abundance --- *B. mendax*, *B. monticola*, and *B. mucidus* --- the lower peak of floral resource abundance is almost entirely outside their elevational range. The tree line, therefore, functions as critical foraging habitat for high-elevation bumble bees. Above the tree line, flora resource availability declined sharply as flower-rich grassland and shrubland transitioned to sparse alpine scree. The observation that the upslope advancement of alpine tree lines often lags behind climate warming [@Dullinger2004-qu] suggests that high elevation bumble bees could move upslope faster than their floral resources, increasingly confined to resource-poor habitat, though the opposite pattern could also occur if the climate-tracking of bumble bees is limited by other factors, such as nest site availability [@Marshall2020-pc]. Moreover, several studies have found that the introduction or removal of bumble bee species from a community can have strong effects on the fitness of other species and on patterns of floral visitation [@Nagamitsu2010-wi; @Brosi2013-mu; @Ishii2013-sm]; if elevational range shifts introduce new competitive pressures via the encroachment of low-elevation species into the historic ranges of high-elevation species, this effect might be expected to be most pronounced at the tree line. The significance of the tree line ecotone as a threshold of floral resource availability and as a zone of overlapping bumble bee distributions corroborates the recent findings of @Minachilis2020-xd in the Mediterranean system of Mount Olympus, indicating that these patterns are likely to generalize well across mountain ecosystems.

The tree line not only punctuated patterns of abundance but also those of species-level and interaction-level β-diversity. While the turnover rate of bumble bee species was low across the whole elevation gradient and showed no response to the tree line, the turnover rate of the floral community was an order of magnitude higher and exhibited a marked inflection point at the tree line, above which it increased steeply. This floristic importance of the tree line has been noted in previous work [@Pellissier2010-kg; @Descombes2017-id], but our study provides a functional extension to this pattern, demonstrating that the tree line also marks a zone of rapid acceleration in the turnover of bumble- bee-flower interactions that is evidently driven by the underlying turnover of the floral community. The picture that emerges is a striking contrast between a relative stable bumble bee community and an extremely dynamic floral community. The dominance that we observed of total interaction β-diversity by floral species turnover is consistent with the findings of @Simanonok2014-ou in the Rocky Mountains of North America, and it is perhaps unsurprising given the exceptional cold-hardiness of bumble bees and the sensitivity of plants to temperature gradients. Nevertheless, it highlights the behavioral challenge that bumble bees have presumably faced for the entirety of their evolutionary history as mountain florivores: how to forage efficiently in a diverse floral community that turns over so rapidly. Indeed, the remarkable flexibility and intelligence of bumble bee foraging [@Heinrich1979-ep; @Loukola2017-bt], which have been interpreted as side-effects of their sociality [@Dukas1991-ig], could be interpreted alternatively as an adaptation to precisely this problem of foraging in the context of extreme β-diversity. A fascinating question that could not be answered by our sampling approach is whether mountain bumble bees adaptively forage up- or downslope, as suggested by @Lundberg1980-qk. Bumble bees have large foraging ranges and have been shown both to track resources through space and time [@Devoto2014-ay] and to cross forest matrix to reach patches of foraging habitat [@Mola2020-yc]. Assuming an average slope of 20° and a foraging range of 1 km, a bumble bee could travel up- or downslope by more than 340 m, thus spanning an elevation belt nearly 700 m wide. Such 3-dimensional foraging would enable bumble bees both to exploit the elevational turnover of floral species and to track preferred species through their elevationally staggered phenology, the latter constituting a sort of physiological time travel [@Van_Straalen1983-df].

The central finding of our study, that elevation gradients are characterized by complex and nonlinear responses of species, communities, and biotic interactions --- punctuated by the tree line --- is almost certainly not an idiosyncrasy of bumble bees and their floral hosts but rather a pattern emerging from fundamental physical constraints that should be expected to obtain broadly across the biotic communities of mountains worldwide. Species interactions along tree line ecotones, therefore, deserve special consideration in research and conservation management, since climate-induced range shifts at tree lines should be expected to have disproportionate ecological consequences relative to range shifts at other elevation bands.

@@

<!-- - What do these findings tell use about climate change and range shifts in mountain ecosystems more generally? -->
<!-- though as a matter of applied conservation it is perhaps doubtful that local interventions could forestall for long the consequences of a process occurring at a global scale. -->
